Robust Estimation, Prediction and Control with \\ Linear Dynamics and Generic Costs

### 2020-02-21 main.tex
Preprint version
Updates from Overleaf

% Use the following line for the initial blind version submitted for review:
\usepackage{icml2020}

% If accepted, instead use the following line for the camera-ready submission:
%\usepackage[accepted]{icml2020}

@@

	\begin{icmlauthorlist}
		\icmlauthor{Edouard Leurent}{seq,val,ren}
		\icmlauthor{Denis Efimov}{val}
		\icmlauthor{Odalric Ambrym-Maillard}{seq}
	\end{icmlauthorlist}

	\icmlaffiliation{seq}{Inria Lille Nord-Europe, Sequel Project, Lille, France}
	\icmlaffiliation{val}{Inria Lille Nord-Europe, Valse Project, Lille, France}
	\icmlaffiliation{ren}{Renault Group, Guyancourt, France}

@@

To address this issue, we turn to the framework of \emph{robust} decision-making \citep{Bental2009,Bertsimas2011,Gorissen2015}: instead of merely considering a point estimate of the dynamics, we build an entire \emph{confidence region} that contains the true dynamics with high probability, and optimise the decisions so as to maximise the worst-case outcome with respect to this uncertainty. Maximin problems are notoriously hard to solve, and all the more in our setting where the reward function $R$ is generic. In particular, this excludes a number of tools from the robust synthesis literature, such as $\cH_\infty$ control \citep[see, e.g.][]{Basar1996} or system-level synthesis \citep{Dean2017,Dean2018}. We turn instead to more general planning algorithms. The continuous state space prevents us from using dynamic programming techniques, so we consider tree-based planning instead. Though there exist works addressing continuous actions \citep{Busoniu2018,Weinstein2012}, we resort to a first approximation and discretise the decision space by adopting a hierarchical control architecture. At each time, the agent can select low-level controller to execute among a finite basis, e.g. choose a short-term goal to reach or a skill to execute such as a driving maneuver. The discretisation induces a suboptimality, but it can be mitigated by diversifying the controller basis. Tree-based planning algorithms are relevant in this context of discrete actions, but are only designed for a single known generative model instead of an entire confidence region for the system dynamics. We adapt them to the robust framework by following a two-steps procedure: we first predict a set of states that can be reached under a planned control sequence with respect to all possible dynamics. This reachability set is used to provide a conservative pessimistic reward to the planning algorithm. The next section aims to clarify and formalize these ideas, and state our contributions.

@@

\paragraph{Linear-quadratic problems} There is a rich body of literature dedicated to the study of linear systems with quadratic costs, and we focus on those providing non-asymptotic results when the system parameters $A, B$ are unknown. In their seminal work \citet{abbasi-yadkori11a}, consider the problem of cumulative regret minimisation. They follow the \emph{Optimism in the Face of Uncertainty} paradigm from the multi-armed bandit literature, that consists in selecting the best possible dynamics within a high-confidence region while enforcing a controllability constraint, and computing the corresponding optimal control in closed-form by solving a Riccati equation. They show that this procedures achieves a $\tilde{\cO}\left(N^{1/2}\right)$ cumulative regret. \citep{Ibrahimi2013,Faradonbeh2017} adapt these results to the case of sparse dynamics. \citet{Ouyang2017,abeille18a} provide the same result by applying Thomson Sampling rather of optimism to select a dynamical model, and rejection sampling to enforce the controllability constraint. These methods aim at regret minimization to quickly reduce uncertainty, through information-seeking behaviours. Other works use noise injection for exploration such as \citep{Dean2017,Dean2018}. However, neither optimism and cumulative regret minimization nor random exploration fit a critical setting, where ensuring safety as a top priority requires instead to consider the pessimistic outcomes.
The work of \citet{Dean2017} is closest to our setting: after an offline estimation phase, where the controls do not aim at regret minimization, they estimate a simple regret between a robust controller and the optimal performance. Our work mainly differs in that it addresses generic costs and forbids random exploration.

\paragraph{Robust planning}
Robust optimization with generic costs has also been studied in the context of finite Markov Decision Processes with uncertain parameters by \citet{Iyengar2005}, \citet{Nilim2005} and \citet{Wiesemann2013}. They showed that the main results of Dynamic Programming can be extended to their robust counterparts only when the dynamics ambiguity set verifies a certain rectangularity property, that amounts to an independence assumption between the uncertainty sets of each transition and is related to time-varying uncertainty. Since our setting requires dealing with continuous states and time-invariant uncertainty, these methods do not apply.

@@

Hence, by using enough computational budget $K$ for planning we can get as close as we want to the optimal surrogate value $\hat{V}^r(a^{\star})$, at a polynomial rate. Unfortunately, there exists a gap between $\hat{V}^r$ and the true robust objective $V^r$, which stems from three approximations: (i) the true reachable set was approximated by an enclosing interval in \eqref{eq:inclusion-property}; (ii) the time-invariance of the dynamics uncertainty $A(\theta)\in\cC_{[N],\delta}$ was handled by the interval predictor \eqref{eq:interval-predictor} as if it were a time-varying uncertainty $A(\theta(t))\in\cC_{[N],\delta},\forall t$ ; and (iii) the lower-bound $\sum\min\leq \min\sum$ used to define the surrogate objective \eqref{eq:surrogate_objective} is not tight. However, this gap can be bounded with additional assumptions.
\begin{theorem}[Regret bound]
\label{thm:control-error}
Under three conditions:
\begin{enumerate}
	\item a Lipschitz regularity assumption for the reward $R$;
	\item a persistent excitation (PE) assumption:
	\begin{align*}
	\exists \underline{\phi},\overline{\phi}>0: \forall n\geq n_0,\quad \underline{\phi}^2 \leq \lambda_{\min}(\Phi_{n}^\transp\Sigma_{p}^{-1}\Phi_{n}) \leq \overline{\phi}^2;
	\end{align*}
	\item a stability condition: there exist $P>0,Q_0\in\Real^{p\times p}$, $\rho>0$ such that
	$\begin{bmatrix}
		A_0^\transp P + P A_0^\transp + Q_0 & P|D|  \\
		|D|^\transp P & -\rho I_r \\
	\end{bmatrix}< 0;$
\end{enumerate}
 we can bound the approximation error of \autoref{alg:full}:
\begin{equation*}
\hat{V}^r(\bu) \leq {V}^r(\bu) \leq V(\bu) \leq \hat{V}^r(\bu) + \Delta_\omega + \cO\left({\log N / N}\right)
\end{equation*}
with probability at least $1-\delta$, where $\Delta_\omega$ is a constant which corresponds to an irreductible suboptimality suffered from being robust to instantaneous perturbations $\omega(t)$.
\end{theorem}

@@

In both experiments, we used $\gamma=0.9$,  $\delta=0.9$ and a planning budget $K=100$. The perturbations were sampled uniformly in $[-0.1, 0.1]^r$ while the measurements are gaussian with covariance $\Sigma_r = 0.1 I_s$.

>>>>

% Use the following line for the initial blind version submitted for review:
%\usepackage{icml2020}

% If accepted, instead use the following line for the camera-ready submission:
\usepackage[accepted,preprint]{icml2020}

@@

	\begin{icmlauthorlist}
		\icmlauthor{Edouard Leurent}{seq,val,ren}
		\icmlauthor{Denis Efimov}{val}
		\icmlauthor{Odalric-Ambrym Maillard}{seq}
	\end{icmlauthorlist}

	\icmlaffiliation{seq}{Inria Lille Nord-Europe, SequeL Project, Lille, France}
	\icmlaffiliation{val}{Inria Lille Nord-Europe, Valse Project, Lille, France}
	\icmlaffiliation{ren}{Renault Group, Guyancourt, France}

@@

To address this issue, we turn to the framework of \emph{robust} decision-making \citep{Bental2009,Bertsimas2011,Gorissen2015}: instead of merely considering a point estimate of the dynamics, we build an entire \emph{confidence region} that contains the true dynamics with high probability, and optimise the decisions so as to maximise the worst-case outcome with respect to this uncertainty. Maximin problems are notoriously hard to solve, and all the more in our setting where the reward function $R$ is generic. In particular, this excludes a number of tools from the robust synthesis literature, such as $\cH_\infty$ control \citep[see, e.g.][]{Basar1996} or system-level synthesis \citep{Dean2017,Dean2018}. We turn instead to more general planning algorithms. The continuous state space prevents us from using dynamic programming techniques, so we consider tree-based planning instead. Though there exist works addressing continuous actions \citep{Busoniu2018,Weinstein2012}, we resort to a first approximation and discretise the decision space by adopting a hierarchical control architecture. At each time, the agent can select low-level controller to execute among a finite basis, e.g. choose a short-term goal to reach or a skill to execute such as a driving maneuver. The discretisation induces a suboptimality, but it can be mitigated by diversifying the controller basis. Tree-based planning algorithms are relevant in this context of discrete actions, but are only designed for a single known generative model instead of an entire confidence region for the system dynamics. We adapt them to the robust framework by following a two-steps procedure: we first predict a set of states that can be reached under a planned control sequence with respect to all possible dynamics. This reachability set is used to provide a conservative pessimistic reward to the planning algorithm. The next section aims to clarify and formalise these ideas, and state our contributions.

@@

\paragraph{Linear-quadratic problems} There is a rich body of literature dedicated to the study of linear systems with quadratic costs, and we focus on those providing non-asymptotic results when the system parameters $A, B$ are unknown. In their seminal work \citet{abbasi-yadkori11a}, consider the problem of cumulative regret minimisation. They follow the \emph{Optimism in the Face of Uncertainty} paradigm from the multi-armed bandit literature, that consists in selecting the best possible dynamics within a high-confidence region while enforcing a controllability constraint, and computing the corresponding optimal control in closed-form by solving a Riccati equation. They show that this procedures achieves a $\tilde{\cO}\left(N^{1/2}\right)$ cumulative regret. \citep{Ibrahimi2013,Faradonbeh2017} adapt these results to the case of sparse dynamics. \citet{Ouyang2017,abeille18a} provide the same result by applying Thomson Sampling rather of optimism to select a dynamical model, and rejection sampling to enforce the controllability constraint. These methods aim at regret minimisation to quickly reduce uncertainty, through information-seeking behaviours. Other works use noise injection for exploration such as \citep{Dean2017,Dean2018}. However, neither optimism and cumulative regret minimisation nor random exploration fit a critical setting, where ensuring safety as a top priority requires instead to consider the pessimistic outcomes.
The work of \citet{Dean2017} is closest to our setting: after an offline estimation phase, where the controls do not aim at regret minimisation, they estimate a simple regret between a robust controller and the optimal performance. Our work mainly differs in that it addresses generic costs and forbids random exploration.

\paragraph{Robust planning}
Robust optimisation with generic costs has also been studied in the context of finite Markov Decision Processes with uncertain parameters by \citet{Iyengar2005}, \citet{Nilim2005} and \citet{Wiesemann2013}. They showed that the main results of Dynamic Programming can be extended to their robust counterparts only when the dynamics ambiguity set verifies a certain rectangularity property, that amounts to an independence assumption between the uncertainty sets of each transition and is related to time-varying uncertainty. Since our setting requires dealing with continuous states and time-invariant uncertainty, these methods do not apply.

@@

Hence, by using enough computational budget $K$ for planning we can get as close as we want to the optimal surrogate value $\hat{V}^r(a^{\star})$, at a polynomial rate. Unfortunately, there exists a gap between $\hat{V}^r$ and the true robust objective $V^r$, which stems from three approximations: (i) the true reachable set was approximated by an enclosing interval in \eqref{eq:inclusion-property}; (ii) the time-invariance of the dynamics uncertainty $A(\theta)\in\cC_{[N],\delta}$ was handled by the interval predictor \eqref{eq:interval-predictor} as if it were a time-varying uncertainty $A(\theta(t))\in\cC_{[N],\delta},\forall t$ ; and (iii) the lower-bound $\sum\min\leq \min\sum$ used to define the surrogate objective \eqref{eq:surrogate_objective} is not tight. However, this gap can be bounded with additional assumptions.
\begin{theorem}[Regret bound]
\label{thm:control-error}
Under three conditions:
\begin{enumerate}
	\item a Lipschitz regularity assumption for the reward $R$;
	\item a persistent excitation (PE) assumption:
	\begin{align*}
	\exists \underline{\phi},\overline{\phi}>0: \forall n\geq n_0,\quad \underline{\phi}^2 \leq \lambda_{\min}(\Phi_{n}^\transp\Sigma_{p}^{-1}\Phi_{n}) \leq \overline{\phi}^2;
	\end{align*}
	\item a stability condition: there exist $P>0,Q_0\in\Real^{p\times p}$, $\rho>0$ such that
	$\begin{bmatrix}
		A_0^\transp P + P A_0^\transp + Q_0 & P|D|  \\
		|D|^\transp P & -\rho I_r \\
	\end{bmatrix}< 0;$
\end{enumerate}
 we can bound the approximation error of \autoref{alg:full}:
\begin{equation*}
\hat{V}^r(\bu) \leq {V}^r(\bu) \leq V(\bu) \leq \hat{V}^r(\bu) + \Delta_\omega + \cO\left({\log N / N}\right)
\end{equation*}
with probability at least $1-\delta$, where $\Delta_\omega$ is a constant which corresponds to an irreducible suboptimality suffered from being robust to instantaneous perturbations $\omega(t)$.
\end{theorem}

@@

In both experiments, we used $\gamma=0.9$,  $\delta=0.9$ and a planning budget $K=100$. The perturbations were sampled uniformly in $[-0.1, 0.1]^r$ while the measurements are Gaussian with covariance $\Sigma_r = 0.1 I_s$.
